---
title: "Plotly Presentation"
author: "Milton Candela"
date: "1/23/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plotly)
library(dplyr)
```

## Introduction

This exercise requires the following libraries:

```{r libraries, eval = FALSE, warning = FALSE, error = FALSE, echo = TRUE}
library(plotly)
library(dplyr)
```

We will also be using the *rock* dataset from the base **datasets** package already installed in R, this is how the dataframe looks:

```{r dataframe}
head(rock)
```

## Plot created by Plotly

Our data set represent different Measurements on Petroleum Rock Samples, we will plot all four elements with a 3D scatter plot, the color represents the permeability in milli-Darcies.

```{r cars, echo = FALSE}

plat <- plot_ly(rock, x = ~area, y = ~peri, z = ~shape, type = 'scatter3d',
                mode = 'markers', color = ~perm) %>%
       layout(scene = list(xaxis = list(title = 'Area of pores (Pixels)'),
                           yaxis = list(title = 'Perimeter (Pixels)'),
                           zaxis = list(title = 'Perimeter/sqrt(area)')))
plat
```